---
title: "R workshop - data visualization"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Participants will be able to visualize their data. This implies basic visualization with fast core functions (plot, hist, boxplot), but also advanced visualization with elaborate formatting by GGPLOT2. Figures will be combined as via GRIDEXTRA and saved to a file. Opionally, we will discuss the animation of figures used for shiny apps or webpages by GGANIMATE. (by Nico Bast)

# helpful resources

example graphs and code: https://www.r-graph-gallery.com/

ggplot reference: https://ggplot2.tidyverse.org/reference/index.html 

# setup

run this before executing the code

## required packages

```{r required_packages}

#check install status of required packages - install if not found
list.of.packages <- c( 'reshape2','ggplot2','gganimate', 'gridExtra')
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

#load required packages 
suppressMessages({
require(reshape2) #data manipulation --> reshape2::melt (melting data to long format)
require(ggplot2) #advanced visualization
require(gridExtra) #combine plots 
require(gganimate) #animater plots
})

```

## load data

load two objects to current environment:

1. data with i=948 participants (rows) and k=45 variables

2. group as group identifier

```{r load_data}

#read data
load('data/efa_data.Rdata')

```


# basic plots 

These visualization methods are built-in with R base packages. They are very fast even with large data sets and are often applied to visualize data as checks on-the-fly (distribution, outlier, etc.).

## plot

PLOT is a GENERIC function - thus object class defines applied method (see "methods(plot)")

```{r basic_visualization_baseplot}

methods(plot) # shows different subfunctions of plot based on data class

#prepare the data
sum_score<-rowSums(data_efa,na.rm=T) #create a sum score

# -- PLOT (scatterplot, barplot, boxplot, etc.)####
plot(sum_score) #SCATTERPLOT: numeric data is plotted across id
plot(group,sum_score) #plot function interprets group as numeric variable and applies a scatter plot (plot.default)
plot(as.factor(group),sum_score)  #plot function interprets group as factor and applies boxplots (plot.factor)
plot(as.factor(group)) # categorical data is plotted as barplot
```

## histogram

histogram function is very fast, even for large datasets.
histogram of investigated variables allows a quick check whether data is suitable for analysis

```{r histogram}

hist(sum_score) #basic histogram

hist(sum_score,
     breaks=30, #number of breaks / number of bars
     col='pink') #color of bars

plot(density(sum_score,na.rm=T)) #density plot, similar to histogram, only useful in bigger datasets

```

## formatting

graphcial parameters can be adjusted by a call withing the function to change appearance of the plot

```{r format_basic plots}

plot(sum_score,
     col=group, #color according to group
     pch=16, #point type
     asp= 3/1, #ratio of y/x axes
     main='Sum scores by participant',ylab = 'raw sum score') #labelling

#you can also add annotations after graph is plotted:
#abline(v=400) 
#abline(h=0)

```

## change parameters

graphical parameters (PAR function) can be called to change parameters for the duration of the session.
This allows to initialize favorite parameters in the setup chunks of the script

### number of plots

```{r change_parameters}

# ?par # options for graphical parameters

par(mfrow=c(1,2)) #MFROW: change the graphical parameter for the number of plots to be drawn (here: one row, two columns)
  ##--> BE CAREFUL: PERMANENT FOR THE SESSION

plot(sum_score,col=group,cex=0.8,pch=16,ylab = 'raw sum score') #scatterplot on left side (called first)
plot(as.factor(group),sum_score,col=1:4,xaxt='n',xlab='group') #boxplot on right side with xaxis removed

  #added graphical parameters in new line (will overplot current plot)
  axis(1, at=1:4, labels=c('ASD','ID','MD','CG')) #add a customized axis
  mtext("Scatterplot and Boxplot of Sum Score", side=3, outer=TRUE, line=-3) #add a custom label for the grit plot

par(mfrow=c(1,1)) #reset graphical parameter 

```

### color palette

```{r color_palette}

palette() #displays the palette of default colors (applied to graphical parameter col)
#colors() #color names that can be applied to palette
palette(c('springgreen3','tomato','turquoise','peachpuff1')) #change default palette to specific colors

plot(as.factor(group),sum_score,col=1:4,xlab='group') #boxplot with new color palette

palette('default') #reset palette to default

```

## save plot to file

base packages allow to save as JPEG, BMP, and TIFF file format. TIFF Format is usually requested by journals

```{r save_basic_figure}

tiff(file="output/figure1_histogram.tiff", # create a file in tiff format in current working directory
width=6, height=4, units="in", res=300) #define size and resolution of the resulting figure
hist(sum_score,breaks=30,col='pink',main='histogram of sumscore') #draw object to file (will not be shown)
dev.off() #close operation and save file

# additional argument - reduces size of Tiff files by compression
# compression = 'lzw' 

```

# advanced plots (ggplot2)

GGPLOT = Grammar of graphics plotting. GGPLOT2 is a powerful package by Hadley Wickham with a standardized syntax to allow for elaborate plotting. There are whole books dealing with the syntax of ggplot2: https://ggplot2-book.org/index.html

## basic syntax

the basic components that needed for every ggplot2 are DATA, AESTHETICS, and GEOM

1. DATA - data that is visualized

2. AESTHETICS (aes) - how the visualized data is structured (corrdinate system)

3. GEOM - visualization type

```{r basic_syntax}
  df.plot<-data.frame(group,sum_score) #DATA - needs to contain all variables to be plotted in one object
  
  g<-ggplot(data=df.plot, #DATA
            aes(x=as.factor(group),y=sum_score)) #AESTHETICS (defines coordinate system)
  
  g+geom_boxplot() #GEOM (defines the type of plot) 
  g+geom_violin() #different GEOM
  
```

## prepare the data

data (data_efa) current is in WIDE unstacked format (xlsx, SPSS standard formats) with in different column based on context.
ggplot requires data to be in LONG format as stacked data with one (single) column for each variable. Other variables may specify the context of an index variable

```{r transform_to_long}

  id<-seq_len(nrow(data_efa)) #add a participant id variable (usually row number in wide data format)
  data_efa<-data.frame(id,group,data_efa) #concatenate to a data.frame
  df_long<-melt(data_efa, id.vars = c('id','group'), measure.vars = 3:47) #melt a wide data frame to LONG FORMAT
  str(df_long) #show structure
  
```

## aggregate data

stat_summary are a specific form of geom that transforms/aggregates the data

```{r, fig.width=10,fig.height=6}
 
  ggplot(data=df_long, #DATA
         aes(x=variable,y=value,group=group,color=as.factor(group)))+ #AES
    
    stat_summary(fun.data = 'mean_se') #GEOM
    
```

## layers

### load new data

```{r load running_data}

load('data/running_data.Rdata')
df.running<-df.running[df.running$duration>0,] #correct for artifacts

```

### inspect data

```{r newdata_basic_plots}

  names(df.running)
    
  par(mfrow=c(2,2))  
    
  with(df.running,hist(distance,breaks=30))
  with(df.running,hist(elevation_gain))
  with(df.running,hist(temperature))
  
  with(df.running,plot(temperature))
  
  par(mfrow=c(1,1))
  
```

### linear model

```{r linear_model}
 with(df.running,summary(lm(distance~hour.class+day.class+year.class+temperature+duration))) #linear model
```

### multiple layers

Different geoms can be combined in one plot. Question: How did running distance develop over the years?

```{r multiple_layers}

  #DEFINE BASIC PLOT (DATA, AESTHETICS)
  g<-ggplot(df.running[order(df.running$run.date),],aes(x=run.date,y=distance/1000)) 
  #ADD ADDITIONAL COMPONENTS:
  g+geom_point(alpha=0.5)+ #GEOM: point
    geom_boxplot(aes(group=year.class),color='orange',fill='wheat')+ #GEOM: boxplot 
    geom_smooth(method='loess',color='navy',fill='blue')+ #GEOM: smoothed line - general additive model
    labs(x=NULL,y='running distance (km)',title='Running distance over the years')+ #LABELS
    theme_bw()+ #THEME - add a predefined overall appearance
    ylim(0,20) #ADDITIONAL GRAPHICAL PARAMETERS - here, limit y-axis to values between 0 and 20

```

## facets

Facets allow to draw multiple plots based on grouping variable. Question: Is running distance affected by weekdays and years?

```{r facets}

  g<-ggplot(df.running,aes(x=as.factor(day.class),y=distance/1000)) 
  g+geom_violin(fill='lightblue')+ #GEOM
      labs(x='weekdays',y='running distance (km)',title='Running distance across weeksdays and years')+ #LABELS
      theme_bw()+ #THEME
      ylim(0,20)+ #ADDITIONAL PARAMETERS (limit y-axis)
      scale_x_discrete(labels=c('Mo','Tu','We','Th','Fr','Sa','Su'))+ #ADDITIONAL PARAMETERS (relabel x axis)
      facet_wrap(~year.class) #FACETS: wrap (one dimensional variable)
  
   g<-ggplot(df.running,aes(x=distance/1000)) 
   g+geom_histogram(bins=20)+
     xlim(0,20)+
     facet_grid(day.class~year.class) #FACETS: grid (two dimensional variables)
  
```

## annotation

plots can be extended by annotations that enrich the plot with additional data that is not part of the drawn dataset

```{r annotation, fig.width=10,fig.height=7}

g<-ggplot(df.running,aes(x=calories))

g+geom_histogram(color='blue',fill='violet')+
   annotate('rect',xmin=1500,xmax=3100,ymin=-10,ymax=10,alpha=0.4,fill='orange')+
   annotate('text',x=2200,y=20,label="longer runs",color='orange')+
   annotate('text',x=1500,y=130,label=paste('Running burns calories! \n median:',median(df.running$calories),'kcal'),
            cex=6,color='blue')+
   geom_vline(xintercept = median(df.running$calories),cex=2,col='blue') +
   geom_hline(yintercept = 100, lty=2) +
    theme_bw()

```

## save to file

R can save plots in common graphics formats (personally prefer tiff: high quality, reasonable size)

```{r save_to_file}

  #create a TIFF FILE in specified path
  tiff(filename = "output/figure2_another_histogram.tiff", 
  width = 6, height = 3, units = "in", pointsize = 12, res=300, compression="lzw") #define dimensions

  #DRAW object to be saved
  g<-ggplot(df.running,aes(x=distance)) 
  g+geom_histogram(aes(y=..density..),fill='royalblue4',color='lightblue')+
    stat_function(fun = dnorm, args = with(df.running, c(mean = mean(distance), sd = sd(distance))))+ #STAT_FUNCTION (applies function to all x values)
    labs(x='distance (m)',title='Running distance of the last 7 years')+
    theme_minimal()
  #figure is SAVED to specified graphics file
  dev.off()
  
```


# multiple plots (grid.arrange)

It may often be useful to combine plots in one figure: https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html

```{r multiple_plots_arrange, fig.width=10,fig.height=7}

g1<-ggplot(df.running,aes(x=calories))+
      geom_histogram(color='blue',fill='violet')+
      annotate('rect',xmin=1500,xmax=3100,ymin=-10,ymax=10,alpha=0.4,fill='orange')+
      annotate('text',x=2200,y=20,label="longer runs",color='orange')+
      annotate('text',x=1500,y=130,label=paste('Running burns calories! \n median:',median(df.running$calories),'kcal'),
            cex=6,color='blue')+
      geom_vline(xintercept = median(df.running$calories),cex=2,col='blue') +
      geom_hline(yintercept = 100, lty=2) +
      theme_bw()

g2<-ggplot(df.running[order(df.running$run.date),],aes(x=run.date,y=distance/1000))+ 
        geom_point(alpha=0.5)+ #GEOM: point
        geom_boxplot(aes(group=year.class),color='orange',fill='wheat')+ #GEOM: boxplot 
        geom_smooth(method='loess',color='navy',fill='blue')+ #GEOM: smoothed line - general additive model
        labs(x=NULL,y='running distance (km)',title='Running distance over the years')+ #LABELS
        theme_bw()+ #THEME - add a predefined overall appearance
        ylim(0,20) #ADDITIONAL GRAPHICAL PARAMETERS - here, limit y-axis to values between 0 and 20

grid.arrange(g1,g2,
             ncol=2, widths=2:1,
             top = "Main Title")


```

